---
title: "Report"
author: "Carlos Espino"
date: "December 12, 2015"
output: pdf_document
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```

\section{Introduction}

\section{Dataset}

```{r, echo=FALSE}
library(ISLR)
library(leaps)
#library(glmnet)
library(rpart.plot)
library(caret)
library(MASS)
library(GGally)
library(grid)
library(ggplot2)
library(gridExtra)
library(splines)
library(caret)
library(knitr)
library(stats)
load("ggthemes_data.rda")

######################################################################
# Setup
######################################################################
source("theme_fivethirtyeight.R")
Crime <- read.csv("Crime.csv")
Crime = Crime[,-1] # Remove the index column

# Identify and remove the outliers found from later analyses
# fit.all <- lm(crmrte~bs(pctmin, knots=pctmin.knots) + log(prbconv) + log(polpc) + prbarr + density:county + prbarr:prbpris + pctmin:polpc + polpc:wfed + density:pctmin + density:pctymle + taxpc:wfed + region:wsta, data=Crime)
bad = c(440,353,437,439)
Crime = Crime[-bad,]
```

The dataset contains a total of 626 observations. 3 outliers were removed according to the result of a residual analysis in a linear regression. We considered them as not representative of the data. 

| county   | county identifier                                                              |   |
|----------|--------------------------------------------------------------------------------|---|
| year     | year from 1981 to 1987                                                         |   |
| crmrte   | crimes committed per person                                                    |   |
| prbarr   | 'probability' of arrest                                                        |   |
| prbconv  | 'probability' of conviction                                                    |   |
| prbpris  | 'probability' of prison sentenc                                                |   |
| avgsen   | average sentence, days                                                         |   |
| polpc    | police per capita                                                              |   |
| density  | people per square mile                                                         |   |
| taxpc    | tax revenue per capita                                                         |   |
| region   | one of 'other', 'west' or 'central'                                            |   |
| smsa     | 'yes' or 'no' if in SMSA                                                       |   |
| pctmin   | percentage minority in 1980 wcon weekly wage in construction                   |   |
| wtuc     | weekly wage in trns, util, commun                                              |   |
| wtrd     | weekly wage in whole sales and retail trade                                    |   |
| wfir     | weekly wage in finance, insurance and real estate                              |   |
| wser     | weekly wage in service industry                                                |   |
| wmfg     | weekly wage in manufacturing                                                   |   |
| wfed     | weekly wage of federal emplyees                                                |   |
| wsta     | weekly wage of state employees                                                 |   |
| wloc     | weekly wage of local governments employees mix offence mix: face-to-face/other |   |
| pctymle  | percentage of young males                                                      |   |

We analized the variables in the dataset starting with the target variable: 'crmrte', the crime rate. Along this study, we will use this variable in diferent forms. We define a categorical value equal to one representing high crime rate, when the value of the target variable is higher that its median value. Also, we will use the natural logarithm of the variable to adequately transform it be able to apply diferent statisticals models to predict and describe the data. The target variable behaviour is represented with a boxplot, a softened histogram and a softened histogram of the logarithm of the variable.

```{r, echo=FALSE}
#WE BUILD A CATEGORICAL VARIABLE WITH THE HIGHEST QUARTILE OF HIGH CRIME RATE
boxplot_crmte_cat = function(variable, data = Crime){
  plot_df= data.frame(y = data[,variable], x = data$crmrte_cat)
  
  plot = ggplot(plot_df) + geom_boxplot(aes(y= y, x = x, fill=x)) + 
        theme_fivethirtyeight() + labs(title = variable, x= "crmrte_cat", y = variable )
  
  return(plot)
}

Crime$crmrte_cat <- rep('low/normal', length(Crime$crmrte))
Crime$crmrte_cat[Crime$crmrte  > median(Crime$crmrte)] <- "high"

ggplot() + geom_boxplot(aes(x=1, y=Crime$crmrte))+ labs(title='boxplot - crime rate')

#histogram of the response variable
ggplot() + geom_density(aes(x = Crime$crmrte), alpha=.3, fill = "grey") + xlab("Crimes per person") +
  theme_fivethirtyeight()

#histogram of the log of the response.
ggplot() + geom_density(aes(x = Crime$crmrte), alpha=.3, fill = "grey") + xlab("log(Crimes per person)") +
  scale_x_log10() +
  theme_fivethirtyeight()
```

Besides the variable target, the dataset contains other 20 variables we used as predictors. Two of them have categorical values. The 'region' variable can have 3 possible values: 'other', 'west' or 'central' and the 'smsa' can have 'yes' or 'no'. The dataset also contains the 'year' variable which can be considered as a time reference. A short description of each variable can be found in the table above. Next, we plot some charts to explore the behaviour of the variables and their relation with the target.

```{r, echo=FALSE}
#we trace paired boxplots for all the variables

grid.arrange(boxplot_crmte_cat(variable = "prbarr"), boxplot_crmte_cat("prbconv"), boxplot_crmte_cat("prbpris"),
          boxplot_crmte_cat("avgsen"), boxplot_crmte_cat("polpc"),boxplot_crmte_cat("density"), ncol=3)

grid.arrange(boxplot_crmte_cat(variable = "taxpc"), boxplot_crmte_cat("pctmin"), boxplot_crmte_cat("wcon"),
             boxplot_crmte_cat("wtuc"), boxplot_crmte_cat("wfir"),boxplot_crmte_cat("wser"), ncol=3)

grid.arrange(boxplot_crmte_cat(variable = "wmfg"), boxplot_crmte_cat("wfed"), boxplot_crmte_cat("wsta"),
             boxplot_crmte_cat("wloc"), boxplot_crmte_cat("mix"),boxplot_crmte_cat("pctymle"), ncol=3)

```

From the boxplot above, we can see that the variables that may have a predictive value with the target are variable 'prbarr', 'density', 'pctmin', 'wfed', 'wmfg' and 'pctymle' as they separate partialy the populations by the value of the target variable defined. In regards to the rest of the predictors, we explore them tracing these folowing charts. We start from the variable 'year'.

```{r, echo=FALSE}
ggplot(Crime) + geom_bar(aes(x =factor(year), fill=crmrte_cat )) +
  scale_fill_fivethirtyeight("cyl") + theme_fivethirtyeight() + xlab("year")
#no significance difference
```

We comment on avobe plot that there is no sifnificance trend on the crime rate along the time line considered. The other two variables with categorical values are 'region' and 'smsa'.

```{r, echo=FALSE}
ggpairs(Crime[,c( 'region','crmrte_cat')])+
  theme_fivethirtyeight()
#candidate to test anova region west

ggpairs(Crime[,c( 'smsa','crmrte_cat')])+
  theme_fivethirtyeight()
#smsa yes with high rate.
```

In these two charts above we see an decrease in the crime rate when the variable 'region' takes value 'west' and when the 'smsa' variable takes the value 'yes'.
In this sense, we continue to explore further the relationtship between these two categorical variables and the target variable by implementing method ANOVA. We fit an analysis of variance model by call to a linear regression for each stratum.

\section{ANOVA models}

In the first analisys, we consider the variable 'region' and agregate their possible values ('west', 'central' and 'other') into ('w' and 'nw'). We obtain the following output.

```{r, echo=FALSE}
Crime$crmrte_cat <- as.factor(Crime$crmrte_cat)
Crime$region_w_nw <- rep('nw', length(Crime$crmrte_cat))
Crime$region_w_nw[Crime$region=='west'] <- 'w'
Crime$region_w_nw <- as.factor(Crime$region_w_nw)

fit.anova <- aov(crmrte ~ region_w_nw, data=Crime)
summary(fit.anova)
```

The result show a very low p-value, which means that the model considerign diferent populations is acurate. We can compare the means of the crime rate between the 'west' region and other regions. The hypotesis null that the median are equals is rejected.

```{r, echo=FALSE}
#compare the means
tapply(Crime$crmrte,Crime$region_w_nw, mean)
```

Now, we repeat the same analisis considering two factors. We include also the other categorical variable: 'smsa'. We fit an ANOVA model and we get the followign output. 

```{r, echo=FALSE}
fit.anova2 <- aov(crmrte ~ region_w_nw + smsa, data=Crime)
summary(fit.anova2)
```

Again we obtain a good p-valio for each of the two variables, which means that both factors have a strong relation with the variable. The hypotesis null that the medians are equals is rejected.
Besides the two categorical variables, we can include in the analysis of the variance the interaction efect between the two variables. 

```{r, echo=FALSE}
#anova using two categorical values and their interaction
fit.anova3 <- aov(crmrte ~ region_w_nw + smsa + region_w_nw:smsa, data=Crime)
summary(fit.anova3)
```

The p-values indicates that the two factors and the interaction between them are significant.

\section{Confidence Interval for the Median}
As discussed above, we defined a target variable in a categorical format ('high', 'low/normal') if the value of the crime rate was higher that the median to be able to run, in next sections, models that require it that way. Therefore, it would be very usefull to have confidence intervals about the median. The median value for the crime rate is calculated as follows.

```{r, echo=FALSE}
median <- median(Crime$crmrte)
median
```

We can simply obtain a confidence interval arround that value. Considering the binomial probabilty with n = 626 observations and a probability of 0.5, we want to obtain the kth than returns the 98%, the 95%, the 88% and the 81% of the probability by doing the following

1- 2*pbinom(k, 626, 0.5)

the kth values that returns that interval are k= 
```{r, echo=FALSE}
k<- 260+c(23,30,35,38) # probabilities 99%; 95%; 88%; 81%;
```
From the vector of sorted values of crime rate, we select the kth elements of the vector and the n-k+1 elements correspondigly to have the four confidence intervals correspondingly. 

```{r, echo=FALSE}
n <- 626
1 - 2 * pbinom(k, n, 1 / 2)
perc <- c(0.99,0.95,0.88,0.81)
col_nam <- c('signif.%', 'low.level', 'up.level')
conf_i <- matrix(0,nrow=4,ncol=3)
for (i in 1:4){
  conf_i[i,] <- c(perc[i],sort(Crime$crmrte)[k[i]],sort(Crime$crmrte)[n-k[i]+1])
}
colnames(conf_i) <- col_nam
kable(conf_i)
```

A more sofisticated method to obtain the confidence interval for the median

```{r, echo=FALSE}

```

```{r, echo=FALSE}

```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
