---
title: "Report"
author: "Carlos Espino"
date: "December 12, 2015"
output: pdf_document
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```

\section{Introduction}

\section{Dataset}

The dataset 

| county   | county identifier                                                              |   |
|----------|--------------------------------------------------------------------------------|---|
| year     | year from 1981 to 1987                                                         |   |
| crmrte   | crimes committed per person                                                    |   |
| prbarr   | 'probability' of arrest                                                        |   |
| prbconv  | 'probability' of conviction                                                    |   |
| prbpris  | 'probability' of prison sentenc                                                |   |
| avgsen   | average sentence, days                                                         |   |
| polpc    | police per capita                                                              |   |
| density  | people per square mile                                                         |   |
| taxpc    | tax revenue per capita                                                         |   |
| region   | one of 'other', 'west' or 'central'                                            |   |
| smsa     | 'yes' or 'no' if in SMSA                                                       |   |
| pctmin   | percentage minority in 1980 wcon weekly wage in construction                   |   |
| wtuc     | weekly wage in trns, util, commun                                              |   |
| wtrd     | weekly wage in whole sales and retail trade                                    |   |
| wfir     | weekly wage in finance, insurance and real estate                              |   |
| wser     | weekly wage in service industry                                                |   |
| wmfg     | weekly wage in manufacturing                                                   |   |
| wfed     | weekly wage of federal emplyees                                                |   |
| wsta     | weekly wage of state employees                                                 |   |
| wloc     | weekly wage of local governments employees mix offence mix: face-to-face/other |   |
| pctymle  | percentage of young males                                                      |   |

```{r}
summary(cars)
```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
